---
title: "Crime in NYC Parks"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = "center", fig.width = 8, fig.height = 6)
library(ggplot2)
library(ggridges)
library(tidyverse)
library(readxl)
library(stringr)
library(purrr)
library(ggridges)
library(kableExtra)
```
# Motivation

Exploring park crimes in NYC is essential for several reasons:  

+ Improving Public Safety: Identifying high-risk areas allows for targeted interventions, ensuring the protection of parkgoers.  

+ Raising Community Awareness: Understanding crime patterns guides public education and prevention efforts to reduce risks in parks.  

+ Addressing Social and Economic Impact: Crime in parks can deter park usage, affecting local businesses and community engagement, which crime analysis can help mitigate.  

+ Targeting Specific Times or Groups: Seasonal and demographic crime patterns allow for targeted safety measures tailored to specific risks.


# Dataset

For this part, we chose [NYC Park Crime Statistics](https://www.nyc.gov/site/nypd/stats/crime-statistics/park-crime-stats.page) provided by NYPD. This dataset contains detailed records of reported crimes in New York City's parks. It includes data on various crime types, such as robbery, assault, and larceny, categorized by park, borough, and time period. This dataset is updated quarterly and aims to enhance public safety by offering transparency on crime trends in recreational areas. It serves as a valuable resource for researchers, policymakers, and residents interested in understanding and addressing crime in NYC parks.

The data processing process involves several steps to import, clean, and filter crime data from various park crime files spanning from 2015 to 2024. Since crime data is stored quarterly, we wrote a function to traverse all data file paths and special files were handled individually. The final dataset includes variables:  

- `park`: park names;  
- `borough`: parks' boroughs;  
- `size_acres`: size in acres;  
- `category`: categories of parks;  
- Crime types: `murder`, `rape`, `robbery`, `felony_assault`, `burglary`, `grand_larceny`, `grand_larceny_of_motor_vehicles`;  
- `total`: total crimes happened in that park;  
- `year`: statistical year;  
- `season`: statistical season.
  
Each row is a crime record for a park in specific year and season.


```{r include=FALSE, message=FALSE, warning=FALSE}
# Define a function to import and clean a single file with exceptions
import_park_crime_data = function(year, season) {
  # Skip 2024 Q4
  if (year == 2024 && season == "q4") {
    return(NULL)
  }
  
  # Determine the file path
  file_path = sprintf("data/nycparks_crime/%d/nyc-park-crime-stats-%s-%d.xlsx", year, season, year)
  
  # Use skip = 4 for specific files, otherwise skip = 3
  skip_rows = ifelse((year == 2015 && season %in% c("q1", "q2", "q3")) || (year == 2018 && season == "q2"), 4, 3)
  
  # Attempt to read the file
  data = tryCatch(
    {
      read_excel(file_path, skip = skip_rows) |>
        janitor::clean_names()
    },
    error = function(e) {
      warning(sprintf("Failed to read file: %s. Error: %s", file_path, e$message))
      return(NULL) # Skip problematic files
    }
  )
  
  # Check if the 'park' column exists, and proceed accordingly
  if (!"park" %in% names(data)) {
    warning(sprintf("Column 'park' not found in file: %s", file_path))
    return(NULL) # Return NULL for this file if 'park' is missing
  }
  
    # List of columns to ensure numeric
  numeric_columns = c(
    "size_acres", "murder", "rape", "robbery", "felony_assault", 
    "burglary", "grand_larceny", "grand_larceny_of_motor_vehicle", "total"
  )
  
  # Add missing numeric columns as NA
  missing_columns = setdiff(numeric_columns, names(data))
  if (length(missing_columns) > 0) {
    data[missing_columns] = NA
  }
  
  # Ensure these columns are numeric
  data = data |>
    mutate(across(all_of(numeric_columns), as.numeric))
  
  # Perform cleaning and filtering
  data |>
    mutate(
      park = str_replace_all(str_to_lower(park), " ", "_"),
      borough = str_replace_all(str_to_lower(borough), " ", "_"),
      category = str_replace_all(str_to_lower(category), " ", "_"),
      year = year,
      season = season
    ) |>
    filter(total > 0)
}

# Generate a data frame with all combinations of years and seasons
years = 2015:2024
seasons = c("q1", "q2", "q3", "q4")
year_season_combinations = expand.grid(year = years, season = seasons)

# Import and combine all the data
all_park_crime_data = year_season_combinations |>
  pmap_dfr(~ import_park_crime_data(..1, ..2))

all_park_crime_data = all_park_crime_data |>
  filter(
    park != "total"
  )
```



# Exploratory Data Analysis

## Crime Counts by Year in NYC Parks
   
```{r echo=FALSE, message=FALSE, warning=FALSE}
# Prepare data for plotting
crime_data_long = all_park_crime_data |>
  select(year, murder, rape, robbery, felony_assault, burglary, grand_larceny, grand_larceny_of_motor_vehicle) |>
  pivot_longer(
    cols = c(murder, rape, robbery, felony_assault, burglary, grand_larceny, grand_larceny_of_motor_vehicle),
    names_to = "crime_type",
    values_to = "count"
  ) |>
  group_by(year, crime_type) |>
  summarize(total_count = sum(count, na.rm = TRUE), .groups = "drop")

crime_data_long$year = as.factor(crime_data_long$year)

# Plot the data
p1 = ggplot(crime_data_long, aes(x = year, y = total_count, color = crime_type, group = crime_type)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  labs(
    title = "Crime Counts by Year in NYC Parks",
    x = "Year",
    y = "Total Count",
    color = "Crime Type"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "top", # Move the legend to the top
    legend.title = element_text(size = 12, face = "bold"), # Style the legend title
    legend.text = element_text(size = 10), # Style the legend text
    legend.margin = margin(t = 5, b = 5), # Adjust spacing around the legend
    legend.key.width = unit(1, "cm") # Adjust the key width for better readability
  )

p1
```

```{r echo=FALSE,message=FALSE}
ggsave("images/report/crime_1.png", p1, width = 8, height = 6, dpi = 300)
```


It can be judged from the plot that serious crimes like grand larceny of motor vehicle, murder, rape and robbery, have a relative stable account and trend. However, crimes like burglary, felony assault and grand larceny (not motor vehicle but maybe jewelry or other valuables), have some fluctuations. Burglary and robbery has some decrease in 2020, largely because COVID-19 exploded and travel restriction was conducted. However, felony assault may show a potential increase due to factors such as sparse crowds or a limited number of offenders, which create an environment conducive to the crime. Psychologically, the perpetrators may feel emboldened, believing that the lack of witnesses reduces the likelihood of being arrested.

## Crime Proportion

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(plotly)

crime_types = all_park_crime_data |>
  select(murder, rape, robbery, felony_assault, burglary, grand_larceny, grand_larceny_of_motor_vehicle) |>
  summarize(across(everything(), sum, na.rm = TRUE)) |>
  pivot_longer(cols = everything(), names_to = "crime_type", values_to = "total_crimes")

# Create a pie chart
p2 = plot_ly(
  data = crime_types,
  labels = ~crime_type,
  values = ~total_crimes,
  type = 'pie',
  textinfo = 'label+percent',
  insidetextorientation = 'radial'
) |>
  layout(
    title = "Proportion of Different Crime Types (2015–2024)",
    showlegend = TRUE
  )

p2

```



From the pie chart, we can figure out that grand larceny (41.1%), robbery (31%) and felony assault (22%) were main crime types during last ten years.  
The observed crime patterns in parks suggest that property-related offenses, particularly Grand Larceny and Robbery, are the most frequent, likely due to the ease of committing such crimes in less monitored areas, such as parks, where large-scale or repeat offenses can occur. Physical crimes like Felony Assault, while present, are less common, possibly due to lower reporting rates or a focus by law enforcement on non-violent offenses in park settings. Violent crimes like Murder and Rape are rare in parks, reflecting their overall low frequency but higher severity, likely influenced by deterrent measures, stronger enforcement, or the generally lower prevalence of such crimes in public spaces. The low incidence of Grand Larceny of Motor Vehicles in parks might indicate the effectiveness of targeted law enforcement efforts or security measures, such as surveillance, to deter such offenses.

## Heatmap of Crime Counts

```{r echo=FALSE, message=FALSE}
crime_heatmap = all_park_crime_data |>
  group_by(year, borough) |>
  mutate(
    year = as.factor(year)
  ) |>
  filter(
    park != "total"
  )|>
  summarize(total_crimes = sum(total, na.rm = TRUE))


p3 = ggplot(crime_heatmap, aes(x = year, y = reorder(borough, total_crimes, sum), fill = total_crimes)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "white", high = "red") +
  labs(
    title = "Heatmap of Crimes by Year and Borough",
    x = "Year",
    y = "Borough",
    fill = "Total Crimes"
  ) +
  theme_minimal()

p3
```

```{r echo=FALSE,message=FALSE}
ggsave("images/report/crime_3.png", p3, width = 8, height = 6, dpi = 300)
```


This is a heatmap of total crimes happened in different boroughs by year. In this heatmap, the higher the number of crimes, the darker the red color. It's pretty obvious that Manhattan has a high crime rate. However, analyzing by years, there is no obvious changing trend.

## Crime Statistics by Borough
```{r echo=FALSE, message=FALSE}
# Pivot data to a longer format for plotting
crime_by_borough = all_park_crime_data |>
    replace_na(list(
    murder = 0,
    rape = 0,
    robbery = 0,
    felony_assault = 0,
    burglary = 0,
    grand_larceny = 0,
    grand_larceny_of_motor_vehicle = 0
  )) |>
  pivot_longer(
    cols = c(murder, rape, robbery, felony_assault, burglary, grand_larceny, grand_larceny_of_motor_vehicle),
    names_to = "crime_type",
    values_to = "crime_count"
  ) |>
  filter(
    borough != "NA"
  ) |>
  group_by(
    crime_type, borough
  ) |>
  summarise(
    crime_count = sum(crime_count)
  ) |>
  ungroup() |>
  mutate(
    borough = fct_reorder(borough, crime_count, .fun = sum, na.rm = TRUE)
  )


# Plot with facet_wrap
ggplot(crime_by_borough, aes(x = borough, y = crime_count, fill = borough)) +
  geom_col(show.legend = TRUE) +
  facet_wrap(~crime_type, scales = "free_y") +
  labs(
    title = "Crime Counts by Type and Borough",
    x = "Borough",
    y = "Crime Count"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_blank(),
    strip.text = element_text(face = "bold", size = 10)
  )
```

This plot compares crime counts by type and borough in New York City, with each panel representing a specific crime type. Manhattan stands out with the highest crime counts for grand larceny and felony assault, while Brooklyn leads in grand larceny of motor vehicle. The Bronx shows notably high counts for robbery and murder, often surpassing other boroughs for these crimes. In contrast, Staten Island and the Brooklyn/Queens area consistently report the lowest crime counts across all categories. Property crimes like grand larceny and felony assault have significantly higher totals compared to violent crimes such as murder and rape, which show smaller totals.

The differences in crime counts across boroughs and crime types may stem from several factors. Manhattan, with its dense population and economic activity, shows higher grand larceny and felony assault rates due to more opportunities for theft and interpersonal conflicts. The Bronx, often associated with economic challenges, exhibits higher rates of violent crimes like murder and robbery. Brooklyn's urban sprawl and higher car ownership contribute to more grand larceny of motor vehicles. Staten Island’s low crime counts may reflect its smaller population or underreporting. Variations in tourism, socioeconomic conditions, policing, and community infrastructure further influence the distribution of crimes.



## Distribution of Crime Counts by Type in Top 3 Parks

```{r echo=FALSE, message=FALSE}
# Calculate the total crimes per park
crime_data_summary = all_park_crime_data |> 
  group_by(park) |>
  filter(
    park != "total"
  ) |>
  summarize(total_crimes = sum(total, na.rm = TRUE)) |> 
  arrange(desc(total_crimes))


# Get the top 3 parks with the most crimes
top_parks = crime_data_summary |> head(3) |> pull(park)

# Filter the original data to include only the top 3 parks
crime_data_top_parks = all_park_crime_data |> 
  filter(park %in% top_parks) |>
  pivot_longer(
    cols = murder:grand_larceny_of_motor_vehicle,
    names_to = "crime_type",
    values_to = "total_count"
  )

knitr::kable(crime_data_summary |>
               head())

```

Above is the top six parks where most crimes occurred. We want to have a look at these parks' crime distribution by type for assisting you to determine whether this is truly a dangerous park or if it simply has incidents of relatively minor crimes. Thus, we selected the three parks with the highest total number of crimes, and then divided them into statistics and visualizations according to the types of crimes. The following is the statistical table and the chart.



```{r echo=FALSE, message=FALSE}
top_3_crime_park = crime_data_top_parks |>
  group_by(
    park, crime_type
  ) |>
  summarise(
    count = sum(total_count)
  ) |>
  pivot_wider(
    names_from = crime_type,
    values_from = count
  ) |>
  replace_na(
    list(murder = 0)
  )

top3_crime_table = top_3_crime_park |>
  kbl(
    col.names = c(
      "Park Name", 
      rep(c("Burglary", "Felony Assault", "Grand Larceny", 
            "Motor Vehicle Theft", "Murder", "Rape", "Robbery"), 1)
    ),
    caption = "Crime Statistics by Park"
  ) |>
  add_header_above(c(" " = 1, "Crimes" = 7)) |>
  kable_styling(
    full_width = FALSE, 
    bootstrap_options = c("striped", "hover", "condensed"),
    position = "center",
    font_size = 14
  ) |>
  column_spec(1, width = "25em") |> 
  column_spec(2:8, extra_css = "padding-right: 20px; padding-left: 20px;") |>
  row_spec(0, bold = TRUE, extra_css = "padding-bottom: 10px;") |> # Add spacing to header row
  kable_styling(latex_options = "hold_position") |> 
  row_spec(1:3, extra_css = "padding-top: 10px; padding-bottom: 10px;")

top3_crime_table

```


```{r echo=FALSE}
# Prepare the data for plotting
top_3_crime_park_long = top_3_crime_park |>
  pivot_longer(
    cols = -park, # All columns except "park"
    names_to = "crime_type",
    values_to = "count"
  ) |>
  filter(
    crime_type != "murder"
  )

# Plot using ggplot2
p4 = ggplot(top_3_crime_park_long, aes(x = park, y = count, fill = park)) +
  geom_bar(stat = "identity", alpha = 0.8, show.legend = TRUE) +
  facet_wrap(~ crime_type, scales = "free_y", ncol = 3) + # Facet by crime type
  theme_minimal() +
  theme(
    text = element_text(size = 12),
    axis.text.x = element_blank(), # Rotate x-axis labels for readability
    axis.text.y = element_text(size = 10),
    plot.title = element_text(size = 16, hjust = 0.5, face = "bold"), # Center and bold title
    strip.text = element_text(size = 8, face = "bold"), # Style facet titles
    panel.spacing = unit(2, "lines"),
    legend.key.size = unit(0.4, "cm"), # Reduce legend key size
    legend.text = element_text(size = 10), # Adjust legend text size
    legend.title = element_text(size = 11, face = "bold"), # Adjust legend title
    legend.position = "top" # Place legend at the top
  ) +
  labs(
    title = "Crime Distribution by Type Across Top Parks",
    x = "Park",
    y = "Crime Count"
  )

p4

```

```{r echo=FALSE,message=FALSE}
ggsave("images/report/crime_4.png", p4, width = 8, height = 6, dpi = 300)
```

From both the table and the plot, there were no murders in any of the three parks, and felonies were relatively rare. It is reasonable to assume that the high number of crimes in these parks is due to the high number of less serious crimes.


The plot highlights crime distribution across three NYC parks: Coney Island Beach & Boardwalk, Flushing Meadows-Corona Park, and Washington Square Park. Grand larceny is the most frequent crime, particularly in Flushing Meadows-Corona Park and Washington Square Park, likely due to high foot traffic and the presence of valuables. Robbery and felony assault are also significant, especially in Flushing Meadows-Corona Park, which may result from its size and varied spaces that create opportunities for such offenses. Burglary rates are minimal, concentrated primarily in Flushing Meadows-Corona Park, likely due to the limited number of vulnerable structures. Grand larceny of motor vehicles is notable only in Flushing Meadows-Corona Park, possibly reflecting its extensive parking areas. Violent crimes like rape are rare, with no mention of murder, aligning with broader trends of low violent crime rates in public spaces due to deterrent measures and enforcement. These patterns reflect park-specific factors such as size, usage, accessibility, and monitoring.
# Discussion
1. Data Quality and Completeness. 

The dataset relies on reported crimes, which may underrepresent actual crime occurrences due to underreporting, especially in parks located in boroughs with lower socioeconomic conditions.

2. Lack of Predictive Analysis. 

Insights into causation or actionable triggers for crime remain limited, restricting the utility of findings for proactive measures.

3. Visualization Limitations.

Although the visualizations are informative, some charts, such as the heatmap, might oversimplify trends by aggregating data at the borough level. This approach could mask intra-borough variations or park-specific patterns.






